Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2021-01-27 manuscript_Huang.tex
revision is done!

{\color{red}On the other hand, there are many values in the lower left corner of the plane showing small values of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ (Fig. \ref{fig:CElogistic}(b, c, e, f)). This region corresponds to regular orbits of low complexities. Along with the choice of embedding parameters, the length of time series and long term accumulations of the computation precisions play important roles in the lower left region of the complexity planes, which will be further illustrated by Figs. \ref{fig:bifurcation} and \ref{fig:transient}. Out of the left and right corners, complexities of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ show consistent trends along with the degree of chaoticity as will be explained later in Fig. \ref{fig:bifurcation}. } In general, we therefore suggest that it is the deterministic transition behavior among patterns that allows an improved distinction from random noise.

{\color{red}In addition, we observe a linear correlation between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig. \ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig. \ref{fig:CElogistic}(f)). However, we note that this linear correlation appears only for large embedding $D = 6$. This linear correlation coincides with the up-rising brach of the maximal complexity curve $\mathcal{C}_{max}$, which is expected by the theory \cite{SMCbook2010}. }

{\color{red} @ Reik: I'm not sure whether to include the following discussion in the main text as suggested by the referee. The discriminatory power of different SCMs have been further compared in Fig. \ref{fig:CHlogCmax}, which characterize the distances to the maximal complexity $\mathcal{C}_{max}$. In particular, for a given $\mathcal{H}$ we compute a normalized complexity $\mathcal{C}^{N} = (\mathcal{C} - \mathcal{C}_{min}) / (\mathcal{C}_{max} - \mathcal{C}_{min})$. In turn, $\mathcal{C}^{N}$ is bounded in the interval of $[0, 1]$ and the alignment along $\mathcal{C}_{max}$ is reflected by large values of $\mathcal{C}^{N}$. The results are shown in Fig. \ref{fig:CHlogCmax}. The dots larger than 1 are numerical artefacts Fig. \ref{fig:CHlogCmax}(c) which are due to  improper choice of embeddings when the Logistic map shows high periodic iterations.    
\begin{figure*}
        \centering
        \includegraphics[width=2\columnwidth]{CompEntropyCNormalized_Logistic.pdf}
\caption{(Color online) Three SCMs which are normalized by $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$  at a given $\mathcal{H}$, for the logistic map. Embedding dimensions are indicated by legends.  \label{fig:CHlogCmax}}
\end{figure*}
}

@@

{\color{red} For fBM, we show another figure of normalized complexity measures by the corresponding $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ as Fig. \ref{fig:CHlogCmax}. As the embedding dimension $D$ is increased, the dots are more aligned to $\mathcal{C}_{max}$ and extend further to the left half of the complexity--entropy plane. 
\begin{figure*}
        \centering
        \includegraphics[width=2\columnwidth]{CompEntropyCNormalized_fBm.pdf}
\caption{(Color online) Three SCMs which are normalized by $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$  at a given $\mathcal{H}$, for the fractional Brownian motion.  \label{fig:CHlogCmax}}
\end{figure*}
}

@@

Another conceptual improvement is found for $\mathcal{H}_E$ and $\mathcal{C}_E$ in the parameter range where the logistic map presents period doubling bifurcations, for example, at $r = 3.544$ (Fig.~\ref{fig:bifurcation}(g, h)). It may be noticed that $\mathcal{H}_O$, $\mathcal{C}_O$, $\mathcal{H}_W$ and $\mathcal{C}_W$ all exhibit non-zero values in this parameter range despite a purely periodic dynamics. Furthermore, there are jumps of all four measures at the points of period doubling bifurcations, which have been already reported in Ref.~\cite{BandtPRL2002}. These jumps however are not desirable since the complexity of the dynamics does not change when $r$ passes any of those points. For the case of the period doubling bifurcation at $r=3.544$ (replacing a period-4 by a period-8 solution), we show in Fig.~\ref{fig:transient} that the observed jump in the entropy and SCM values may be explained by accumulated numerical errors during the iterations attracted to the periodic-4 points, which yields long transients. For estimations based on time series of finite length as used in this work, $\mathcal{H}_E$ and $\mathcal{C}_E$ are however not affected in the same way. Certainly, we should not over-interpret their capabilities since numerical inaccuracy would be accumulated in a longer iterative process such that different ordinal patterns are identified. In addition, we pass through many periodic windows of different periods as the control parameter $r$ is varied, which prevents us from using a unique predefined number of iterations as initial transients for time series of different period length. Notably, similar jumps have also been observed when the system bifurcates from a period-3 to a period-6 solution at $r = 3.842$.

@@

Some open problems also common to other nonlinear time series analysis methods remain to be further addressed, for example, the dependence of reliable estimates of SCMs on the available time series length. {\color{red}For reliable estimations of traditional entropy $\mathcal{H}_O$ and SCM $\mathcal{C}_O$, the length $N > 5 D!$ has been largely suggested \cite{BandtPRL2002,AmigoPTRSA2014}. However, even longer length is required for the newly proposed SCMs based on transition frequencies because the transition matrix has the dimension $D!\times (D! - 1)$ if self loops are excluded. This should not be a big problem when working with time series from numerical models provided with the rapid increased computing power nowadays. } For time-discrete dynamical systems like chaotic maps, one may easily use longer data series, which however does not seem to improve the estimates very much due to the accumulation of effects due to a limited numerical precision (Fig.~\ref{fig:transient}). {\color{red} The dependence of the new SCMs on the length $N$ becomes crucial for experimental time series studies. One of more interesting problems is to understand the extra information of transition probability based complexity measures since linear correlations have been found for $\mathcal{C}_{E}$ and small entropy $H$, which are expected for a large number of patterns, i.e., $N = D!$ for high embedding $D$ or equivalently $N = D! \times (D!-1)$ transitions. } For continuous systems, together with the embedding parameters, the sampling frequency is another important characteristic which needs to be further explored. Moreover, the generalization of our approach from univariate to bivariate time series analysis shall be further discussed. We noticed that this aspect has remained a largely untouched subject for statistical complexity analysis so far, while it presents an interesting task to show the dependence of complexity on interaction. {\color{red} We may integrate some of the traditional network topological statistical measures of higher order into the SCMs, for instance, clustering coefficients or path lengths, which have been widely discussed to characterize the dynamical properties of the system generating the time series at hand \cite{ZouPR2018}.} We outline further work on the aforementioned topics as subjects of future research.

>>>>

{\color{red}On the other hand, there are many values in the lower left corner of the plane showing small values of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ (Fig. \ref{fig:CElogistic}(b, c, e, f)). This region corresponds to regular orbits of low complexities. Along with the choice of embedding parameters, the length of time series and long term accumulations of the computation precisions play important roles in the lower left region of the complexity planes, which will be further further discussed in Sec. \ref{sec:transi}. Out of the left and right corners, complexities of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ show consistent trends along with the degree of chaoticity as will be explained later in Sec. \ref{sec:transi}.} In general, we therefore suggest that it is the deterministic transition behavior among patterns that allows an improved distinction from random noise.

{\color{red}In addition, we observe a linear correlation between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig. \ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig. \ref{fig:CElogistic}(f)). However, we note that this linear correlation appears only for large embedding $D = 6$ and coincides with the up-rising brach of the maximal complexity curve $\mathcal{C}_{max}$, which is expected by the theory \cite{SMCbook2010}.}

{\color{red} The discriminatory power of different SCMs have been further compared in Fig. \ref{fig:CHlogCmax}, which characterize how well the alignments are to the maximal complexity $\mathcal{C}_{max}$. In particular, for a given $\mathcal{H}$ we compute a normalized complexity $\mathcal{C}^{N} = (\mathcal{C} - \mathcal{C}_{min}) / (\mathcal{C}_{max} - \mathcal{C}_{min})$. In consequence, $\mathcal{C}^{N}$ is bounded in the interval of $[0, 1]$ and the alignments along $\mathcal{C}_{max}$ are reflected by large values of $\mathcal{C}^{N}$. The results are shown in Fig. \ref{fig:CHlogCmax}. For the traditional $\mathcal{C}_{O}$, again we find there are many points in the right half plane (Fig. \ref{fig:CHlogCmax}(a) for all embedding dimensions. In contrast,  there are no points in the region of $\mathcal{H} >0.8$ for $\mathcal{C}_W$ and $\mathcal{C}_E$ (Fig. \ref{fig:CHlogCmax}(b,c)). Note that in Fig. \ref{fig:CHlogCmax}(c), the dots larger than 1 are numerical artefacts when the control parameters $r$ are close bifurcation points resulting in long transients (as will be explained by Fig. \ref{fig:transient} in Sec. \ref{sec:transi}).  
\begin{figure*}
        \centering
        \includegraphics[width=2\columnwidth]{CompEntropyCNormalized_Logistic.pdf}
\caption{(Color online) Normalized complexity -- entropy planes of three SCMs for the logistic map, which are normalized by $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$  at a given $\mathcal{H}$. (a) $\mathcal{C}_O$, (b) $\mathcal{C}_W$, and (c) $\mathcal{C}_E$. In (c), the dots larger than 1 are numerical artefacts when the control parameters $r$ are close bifurcation points resulting in long transients. Embedding dimensions are indicated by legends.  \label{fig:CHlogCmax}}
\end{figure*}
}

@@

{\color{red} For fBM, we show the normalized complexity measures which are obtained by the corresponding $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ as Fig. \ref{fig:CHfbmCmax}. As the embedding dimension $D$ is increased, the dots are more aligned to $\mathcal{C}_{max}$ and extend further to the left half of the complexity--entropy plane. 
\begin{figure*}
        \centering
        \includegraphics[width=2\columnwidth]{CompEntropyCNormalized_fBm.pdf}
\caption{(Color online) Caption is similar with Fig. \ref{fig:CHlogCmax} for normalized complexity -- entropy planes of three SCMs for the fractional Brownian motion.  \label{fig:CHfbmCmax}}
\end{figure*}
}

@@

Another conceptual improvement is found for $\mathcal{H}_E$ and $\mathcal{C}_E$ in the parameter range where the logistic map presents period doubling bifurcations, for example, at $r = 3.544$ (Fig.~\ref{fig:bifurcation}(g, h)). It may be noticed that $\mathcal{H}_O$, $\mathcal{C}_O$, $\mathcal{H}_W$ and $\mathcal{C}_W$ all exhibit non-zero values in this parameter range despite a purely periodic dynamics. Furthermore, there are jumps of all four measures at the points of period doubling bifurcations, which have been already reported in Ref.~\cite{BandtPRL2002}. These jumps however are not desirable since the complexity of the dynamics does not change when $r$ passes any of those points. For the case of the period doubling bifurcation at $r=3.544$ (replacing a period-4 by a period-8 solution), we show in Fig.~\ref{fig:transient} that the observed jump in the entropy and SCM values may be explained by accumulated numerical errors during the iterations attracted to the periodic-4 points, which yields long transients. For estimations based on time series of finite length as used in this work, $\mathcal{H}_E$ and $\mathcal{C}_E$ are however not affected in the same way showing much smaller values (Fig.~\ref{fig:bifurcation}(g, h)). Certainly, we should not over-interpret their capabilities since numerical inaccuracy would be accumulated in a longer iterative process such that different ordinal patterns are identified. In addition, we pass through many periodic windows of different periods as the control parameter $r$ is varied, which prevents us from using a unique predefined number of iterations as initial transients for time series of different period length. Notably, similar jumps have also been observed when the system bifurcates from a period-3 to a period-6 solution at $r = 3.842$.

@@

Some open problems also com	mon to other nonlinear time series analysis methods remain to be further addressed, for example, the dependence of reliable estimates of SCMs on the available time series length. {\color{red}For reliable estimations of traditional entropy $\mathcal{H}_O$ and SCM $\mathcal{C}_O$, the length $N > 5 D!$ has been largely suggested \cite{BandtPRL2002,AmigoPTRSA2014}. However, even longer length is required for the newly proposed SCMs based on transition frequencies because the transition matrix has the dimension $D!\times (D! - 1)$ if self loops are excluded. This should not be a big problem when working with time series from numerical models provided with the rapid increased computing power nowadays. } For time-discrete dynamical systems like chaotic maps, one may easily use longer data series, which however does not seem to improve the estimates very much due to the accumulation of effects due to a limited numerical precision (Fig.~\ref{fig:transient}). {\color{red} The dependence of the new SCMs on the length $N$ becomes crucial for experimental time series studies. One of more interesting problems is to understand the extra information of transition probability based complexity measures since linear correlations have been found for $\mathcal{C}_{E}$ and small entropy $H$, which are expected for a large number of patterns, i.e., $N = D!$ for high embedding $D$ or equivalently $N = D! \times (D!-1)$ transitions. } For continuous systems, together with the embedding parameters, the sampling frequency is another important characteristic which needs to be further explored. Moreover, the generalization of our approach from univariate to bivariate time series analysis shall be further discussed. We noticed that this aspect has remained a largely untouched subject for statistical complexity analysis so far, while it presents an interesting task to show the dependence of complexity on interaction. {\color{red} We may integrate some of the traditional network topological statistical measures of higher order into the SCMs, for instance, clustering coefficients or path lengths, which have been widely discussed to characterize the dynamical properties of the system generating the time series at hand \cite{ZouPR2018}.} We outline further work on the aforementioned topics as subjects of future research.
